Software Architectures for Integrating Workflow Engines in Science Gateways

### 2016-05-13 paper.tex
done fornow with sec 2

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Task encapsulation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow import. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}
\begin{table*}
\centering
\footnotesize
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{Systems} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          Catania Science Gateway Framework~\cite{ardizzone2012decide},
          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
            Apache Airvata~\cite{marru2011apache}, Neurosciece Gateway~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{kruger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
            Science gateways in~\cite{kacsuk2014science}$^w$
            } \\
  Task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
           }\\
  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
\end{tabular}
\caption{Classification of science gateways based on the architecture used to integrate workflow engines. $^c$: based on the Catania Science Gateway Framework. $^w$: based on WS-PGRADE/gUSE. $^n$: used for neuroimaging.}
\label{table:system-classification}
\end{table*}

@@

Nested workflows are abstract architectural patterns that can be
instantiated in the various architectures described previously. We
focus on instantiation with the service invocation model
(Figure~\subref*{archi:nested}-Right) as this is the most used
architecture. In the instantiation, we assume that the parent and
child workflow engines are distinct pieces of software that require
different workflow services invoked by distinct \texttt{d}
interactions. If this is not the case then workflow services can be
collapsed in a single one with a \texttt{d} interaction with
itself. An example of such interaction is the use of hierarchical
workflows in Pegasus~\cite{Deelman201517}.
Workflow engines communicate with infrastructures using
\texttt{b} and \texttt{c$_2$}. Science gateway transfers user data to
infrastructures using \texttt{c$_1$} interactions.

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They are
also used implicitly in several platforms where workflow engines are
wrapped in workflow tasks as any other command-line tool. Nested
workflows were notably used by the SHIWA Science Gateway to implement
so-called Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
various workflow engines in a consistent
platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
used in the SHIWA Simulation Platform for nested workflow execution
with service invocation.
\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/shiwa-science-gateway.pdf}
\caption{Nested workflow execution through SHIWA Science Gateway. The
  parent workflow engine is \texttt{WS-PGRADE}, invoked as a service
  in the Science Gateway (step \texttt{1}, interaction \texttt{d} on
  Figure~\ref{fig:architectures}(V)-Right). Ten different child engines can be used by nested
  workflows, invoked through the \texttt{Submission service} (step
  \texttt{2}, interaction \texttt{d}). Each of these engines can
  submit tasks and transfer data to a distributed computing
  infrastructure (\texttt{DCI}, step \texttt{3}, interactions
  \texttt{b} and \texttt{c$_2$}). User data interactions (\texttt{c$_1$}) and
  application porting ones (\texttt{a}) are not represented. Figure
  reproduced from~\cite{terstyanszky2014enabling} with permission of
  the first author.}
\label{fig:shiwa-architecture}
\end{figure*}

@@

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. Workflows are integrated in the science gateway through
format conversion from a native format to the science gateway
format. Workflow import is usually an offline process that is not
involved in workflow execution. A few systems have implemented
workflow import. In the SHIWA Simulation Platform, it is implemented
through the IWIR language that provides a common language for
portability across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013} and allows to convert
among $n$ workflow languages using $2n$ interactions instead of $n^2$.
Tavaxy~\cite{Abouelhoda2012}
allows to import, merge and run Taverna~\cite{oinn2004taverna} and
Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
cannot be imported, workflows are run by Tavaxy as nested workflows
using their native engine. The work in~\cite{delaGarza2016} describes
workflow import from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and
from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.
\closednote{Tristan}{Add description of a real system}{Tristan}{Would
  require to reproduce a figure from another paper. The concept seems
  simple enough to not require a figure.}

@@

The pool architecture is overall the most scalable, which is not
surprising since it is designed precisely for scalability. Tight
integration is the least scalable and all the other architectures
perform the same overall. The global scalability score, however,
should not conceal \closednote{Naj}{what do you mean by
  mask?}{Tristan}{I guess 'hide' or 'conceal' are better
  words. Fixed.} the unique characteristics of architectures regarding
this criterion. Nested workflows are the only architecture that can
easily accommodate distributed workflow execution, which can be
critical in some cases. At the same time, the scheduling constraints
created by task encapsulation and nested workflows may become problematic \closednote{Naj}{please don't use these from-film-marketing
  words?! :-) }{Tristan}{OK!} depending on the type of targeted
infrastructure. Non-reliable infrastructures\closednote{Naj}{you mean by
  the showstoppers, i.e. the nested workflows which can accommodate
  distributed executions but are limited by scheduling constraints?}{Tristan}{Yes. I rephrased a bit this part.},
for example, could hardly cope with workflow engines being wrapped in
computing tasks as done in task encapsulation. The availability of
multiple engine instances could also become a critical feature for
science gateways with important workloads, which would favor task
encapsulation and pool. \note{Naj}{I feel this paragraph is a bit
  vague, and doesn't distill and contextualize the thorough work
  described before.}

>>>>

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Task encapsulation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow import. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}
\begin{table*}
\centering
\footnotesize
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{Systems} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          Catania Science Gateway Framework~\cite{ardizzone2012decide},
          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
            Apache Airvata~\cite{marru2011apache}, Neuroscience Gateway~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{kruger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
            Science gateways in~\cite{kacsuk2014science}$^w$
            } \\
  Task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
           }\\
  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
\end{tabular}
\caption{Classification of science gateways based on the architecture used to integrate workflow engines. $^c$: based on the Catania Science Gateway Framework. $^w$: based on WS-PGRADE/gUSE. $^n$: used for neuroimaging.}
\label{table:system-classification}
\end{table*}

@@

Nested workflows are abstract architectural patterns that can be
instantiated in the various architectures described previously. We
focus on instantiation with the service invocation model
(Figure~\subref*{archi:nested}-Right) as this is the most used
architecture. In the instantiation, we assume that the parent and
child workflow engines are distinct pieces of software that require
different workflow services invoked by distinct \texttt{d}
interactions. If this is not the case, then workflow services can be
collapsed into a single one with a \texttt{d} interaction with
itself. An example of such interaction is the use of hierarchical
workflows in Pegasus~\cite{Deelman201517}.
Workflow engines communicate with infrastructures using
\texttt{b} and \texttt{c$_2$}. Science gateway transfers user data to
infrastructures using \texttt{c$_1$} interactions.

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They are
also used implicitly in several platforms where workflow engines are
wrapped in workflow tasks as any other command-line tool. Nested
workflows were notably used by the SHIWA Science Gateway to implement
so-called Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
various workflow engines in a consistent
platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
used in the SHIWA Simulation Platform for nested workflow execution
with service invocation.
\begin{figure*}
\centering
\includegraphics[width=1.5\columnwidth]{figures/shiwa-science-gateway.pdf}
\caption{Nested workflow execution through the SHIWA Science Gateway. The
  parent workflow engine is \texttt{WS-PGRADE/gUSE}, invoked as a service
  in the Science Gateway (step \texttt{1}, interaction \texttt{d} on
  Figure~\ref{fig:architectures}(V)-Right). Ten different child engines can be used by nested
  workflows, invoked through the \texttt{Submission service} (step
  \texttt{2}, interaction \texttt{d}). Each of these engines can
  submit tasks and transfer data to a distributed computing
  infrastructure (\texttt{DCI}, step \texttt{3}, interactions
  \texttt{b} and \texttt{c$_2$}). User data interactions (\texttt{c$_1$}) and
  application porting ones (\texttt{a}) are not represented. Figure
  reproduced from~\cite{terstyanszky2014enabling} with permission of
  the first author.}
\label{fig:shiwa-architecture}
\end{figure*}

@@

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. Workflows are integrated in the science gateway through
format conversion from a native format to the science gateway
format. Workflow import is usually an offline process that is not
involved in workflow execution. A few systems have implemented
workflow import. In the SHIWA Simulation Platform, it is implemented
through the IWIR language, which provides a common language for
portability across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013} and allows conversion
among $n$ workflow languages using $2n$ interactions instead of $n^2$.
Tavaxy~\cite{Abouelhoda2012}
enables the import, merging and execution of Taverna~\cite{oinn2004taverna} and
Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
cannot be imported, workflows are run by Tavaxy as nested workflows
using their native engine. The work in~\cite{delaGarza2016} describes
workflow import from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and
from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.
\closednote{Tristan}{Add description of a real system}{Tristan}{Would
  require to reproduce a figure from another paper. The concept seems
  simple enough to not require a figure.}

@@

The pool architecture is overall the most scalable, which is not
surprising since it is designed precisely for scalability. Tight
integration is the least scalable and all the other architectures
perform the same overall. The global scalability score, however,
should not conceal \closednote{Naj}{what do you mean by
  mask?}{Tristan}{I guess 'hide' or 'conceal' are better
  words. Fixed.} the unique characteristics of architectures regarding
this criterion. Nested workflows are the only architecture that can
easily accommodate distributed workflow execution, which can be
critical in some cases. At the same time, the scheduling constraints
created by task encapsulation and nested workflows may become problematic \closednote{Naj}{please don't use these from-film-marketing
  words?! :-) }{Tristan}{OK!} depending on the type of targeted
infrastructure. Non-reliable infrastructures \closednote{Naj}{you mean by
  the showstoppers, i.e. the nested workflows which can accommodate
  distributed executions but are limited by scheduling constraints?}{Tristan}{Yes. I rephrased a bit this part.} ,
for example, could hardly cope with workflow engines being wrapped in
computing tasks as done in task encapsulation. The availability of
multiple engine instances could also become a critical feature for
science gateways with important workloads, which would favor task
encapsulation and pool. \note{Naj}{I feel this paragraph is a bit
  vague, and doesn't distill and contextualize the thorough work
  described before.}
